---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}

library(dplyr)
library(ggplot2)

activity <- read.table("activity.csv", header = TRUE, sep = ",", 
                       na.strings = "NA", stringsAsFactors = FALSE)
activity$date <- as.Date(activity$date)

total.steps.per.day <- summarise(group_by(activity, date), 
                                        total.steps = sum(steps))

str(activity)
```

## What is mean total number of steps taken per day?

```{r}
g <- ggplot(data = total.steps.per.day, aes(total.steps)) 
g <- g + geom_histogram()
g + ggtitle("Total number of steps taken each day")
```

The **mean** total number of steps taken per day:

```{r} 
mean(total.steps.per.day$total.steps, na.rm = TRUE)
```

The **median** total number of steps taken per day:

```{r} 
median(total.steps.per.day$total.steps, na.rm = TRUE)
```

## What is the average daily activity pattern?

```{r}
average.steps <- summarise(group_by(activity, interval),
                                  average = mean(steps, na.rm = TRUE))

g <- ggplot(average.steps, aes(x = interval, y = average))
g <- g + geom_line()
g <- g + xlab("Interval") + ylab("Average number of Steps")
g + ggtitle("Time series plot")
```

```{r}
ind <- which(average.steps$average == max(average.steps$average, na.rm = TRUE))
itv <- average.steps$interval[ind]
```

Maximum number of steps at the `r itv`-th interval.

## Imputing missing values

```{r}
summary(activity$steps)
```

There are `r summary(activity$steps)[[7]]` missing values in the data set.


```{r}
filled.activity <- activity

na.indices <- which(is.na(filled.activity$steps))
na.intervals <- filled.activity$interval[na.indices]

temp <- sapply(na.intervals, 
               function(iterv) {
                       average.steps[average.steps$interval == iterv, "average"]
                       })


filled.activity$steps[na.indices] <- round(unlist(cbind(temp)))


```

Use an integer value (approximated to the mean for that 5-minute interval) to impute missing values.

```{r}

updated.total.steps.per.day <- summarise(group_by(filled.activity, date), 
                                                total.steps = sum(steps))

g <- ggplot(data = updated.total.steps.per.day, aes(total.steps)) 
g <- g + geom_histogram()
g + ggtitle("Total number of steps taken each day in update data set")

```

Updated **mean** total number of steps taken per day:

```{r}

mean(updated.total.steps.per.day$total.steps, na.rm = TRUE)

```


Updated **median** total number of steps taken per day:

```{r}

median(updated.total.steps.per.day$total.steps, na.rm = TRUE)

```

```{r}

updated.total.steps.per.day <- summarise(group_by(filled.activity, date), 
                                                total.steps = sum(steps))

cols <- c("Imputed" = "red", "Original" = "blue")

g <- ggplot(total.steps.per.day, aes(total.steps)) 
g <- g + geom_histogram(data = updated.total.steps.per.day, 
                        aes(colour = "Imputed", fill = "Imputed", alpha = "Imputed"))
g <- g + geom_histogram(data = total.steps.per.day, 
                        aes(colour = "Original", fill = "Original", alpha = "Original"))

g <- g + ggtitle("Total number of steps taken each day for Old vs. New Data set") 
g <- g + scale_color_manual(name="Bar",values=cols) + scale_fill_manual(name="Bar",values=cols) 
g + scale_alpha_manual(name="Bar",values=c("Imputed"=0.2, "Original"=0.2))
```

The impact of imputing missing data can effectively make the estimates of the total daily number of steps since the mean and median are approximated to the old one, as show as in the two histograms.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
filled.activity$type <- "weekday"
filled.activity$type[weekdays(filled.activity$date) %in% c("Saturday", "Sunday")] <- "weekend"

head(filled.activity)
```

```{r}

updated.average.steps <- summarise(group_by(filled.activity, interval, type),
                                          average = mean(steps))

g <- ggplot(updated.average.steps, aes(x = interval, y = average))
g <- g + geom_line(aes(col = type)) + facet_grid(type ~ .)
g <- g + xlab("Interval") + ylab("Average number of Steps")
g + ggtitle("Time series plot for Weekday vs. Weekend")
```
